---
title: "Intro to `ggplot2` & `tidyr`"
subtitle: "Visualizing Home Sales Data, Garfield County, CO"
author: "Casey Bates"
date: "1/31/2019"
output:
  ioslides_presentation: 
    smaller: true
  slidy_presentation: default
  beamer_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 6, fig.height = 3.7)
```

## Data

A single-family home sales dataset and a condo/townhome dataset for Garfield County, Colorado, that contain 2 years of data from summer 2014 through summer 2016. The data is publically available on the [Garfield County Assessor website](https://www.garfield-county.com/assessor/comparable-sales.aspx). 

```{r data_processing, message=FALSE, include=FALSE}
# This chunk includes the data processing steps to wrangle and clean the data
library(tidyverse) # For reading-in, wrangling, visualizing the data, and for pipe operator
library(readxl) # Tidyverse package for reading in Excel files
library(magrittr) # For the '%<>%' pipe operator
library(knitr) # For nicely formatted tables

# Read-in townhomes and condo sales data
townhomes <- read_xlsx("2017-comparable-sales-condos-townhomes.xlsx") 

# Read-in single family home sales data
single_family <- read_xlsx("2017-comparable-sales-single-family.xlsx") 

# Check column names and formats
glimpse(townhomes) 
glimpse(single_family)
# 'sale_price' column was coerced to numeric and special characters '$' and ',' removed

# Replace spaces in column names with underscores ('_') and make names lowercase
# the package 'magrittr' is used for the pipe operator, this is loaded with the 'tidyverse' library
colnames(townhomes) %<>% str_replace_all("\\s", "_") %<>% tolower() 
colnames(single_family) %<>% str_replace_all("\\s", "_") %<>% tolower()

# Rename column "actual_year_built" as "year_built" for consistency between datasets
townhomes <- townhomes %>% 
  rename(year_built = actual_year_built) 

# Add a "classifcation" column to the single_family dataset and set all values to "Single Family"
single_family <- single_family %>% 
  mutate(classification = "Single Family")

# bind_rows() throws an error b/c 'reception' column is numeric in townhomes dataset; convert to character
townhomes$reception <- as.character(townhomes$reception)

# Combine the datasets into one
home_sales <- bind_rows(single_family, townhomes)
glimpse(home_sales)
# No new columns were created during the 'bind_rows()' process, indicating all colums align
# Remove single_family and townhomes dataframes because they are no longer needed
rm(single_family, townhomes)

# unique(home_sales$classification) reveals a "Garage Only" type. Drop this.
home_sales <- home_sales %>% filter(classification != "Garage Only") 

# Rename the column "heated area" to "square_feet" for clarity, though technically not correct
home_sales <- home_sales %>% 
  rename(square_feet = heated_area)
```

## Processing the data

* Import two Excel files: 

  1. single family home sales, and 
  1. condo & townhome sales
* Replace spaces in column names with underscore and make lowercase
* Rename some columns
* Add `classification` column to `single_family` dataset
    * Set all values to "Single Family"
* Use `bind_rows()` to combine the datasets into one
* Remove "Garage Only" observations


## Glimpse of the data
```{r glimpse}
glimpse(home_sales)
```

## `ggplot2` package in R
* Created by Hadley Wickham
* Built on the "Grammar of Graphics" principles
* Core **tidyverse** package
* Every ggplot2 plot has 3 key components:
  * **Data**
  * **Aesthetic mappings** between variables and visuals 
  * Layer(s) to describe how to render each observation (usually created with a **geom** function)

## Basic scatterplot
```{r scatter_location}
ggplot(data = home_sales, aes(x = square_feet, y = sale_price)) +
  geom_point()
```

## Transparency and color by location
```{r scatter_location_color, warning= FALSE}
ggplot(data = home_sales, aes(x = square_feet, y = sale_price, color = location)) +
  geom_point(alpha = 0.5)
```

## Zooming into sales above $1M
```{r scatter_zoom, warning= FALSE}
ggplot(data = home_sales, aes(x = square_feet, y = sale_price, color = location)) +
  geom_point(alpha = 0.5) +
  coord_cartesian(ylim = c(1000000, 3000000))
```

## Add random noise with jitter
```{r scatter_zoom_jitter, warning= FALSE}
ggplot(data = home_sales, aes(x = square_feet, y = sale_price, color = location)) +
  geom_jitter(alpha = 0.5, width = 300, height = 20000) +
  coord_cartesian(ylim = c(1000000, 3000000))
```

```{r, include = FALSE}
# High sale price observations
home_sales %>% arrange(desc(sale_price)) %>% select(c("sale_price", "location", "classification", "bedrooms", "square_feet")) %>% head(30)
```

```{r, include = FALSE}
# Remove erroneous high sale price observations
home_sales_fix <- home_sales %>% filter(location != "RIFLE" | classification != "Townhome" | sale_price != 2200000)
home_sales_fix %>% filter(location == "RIFLE" & classification == "Townhome") %>% arrange(desc(sale_price)) %>% head(10)
```

## Linear model: sale price vs. square ft.
```{r lm_overall_bw, warning= FALSE}
ggplot(data = home_sales, aes(x = square_feet, y = sale_price)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) # Method set to lm for the linear model
```

## Linear model: sale price vs. square ft.
```{r lm_overall_fixbw, warning= FALSE}
ggplot(data = home_sales_fix, aes(x = square_feet, y = sale_price)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) # Method set to lm for the linear model
```

## Linear model: sale price vs. square feet
```{r lm_overall, warning= FALSE}
ggplot(data = home_sales, aes(x = square_feet, y = sale_price, color = location)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) # Method set to lm for the linear model
```

## Linear model with errors removed
```{r lm_overall_fix, warning= FALSE}
ggplot(data = home_sales_fix, aes(x = square_feet, y = sale_price, color = location)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) # Method set to lm for the linear model
```

## Linear model facetted by location
```{r lm_facet, warning= FALSE}
ggplot(home_sales, aes(square_feet, sale_price, color = location)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ location)
```

## Linear model facetted by location
```{r lm_facet_fix, warning= FALSE}
ggplot(home_sales_fix, aes(square_feet, sale_price, color = location)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ location)
```

```{r seattle_rain, include = FALSE}
# Import seattle rain gauge dataset
seattle_rain <- read_csv("Observed_Monthly_Rain_Gauge_Accumulations_Oct_2002_to_May_2017.csv")
```

## Tidying wide datasets
```{r head_rain}
seattle_rain # Display Monthly Rain Gauge Accumulations for Seattle
```

## `gather` columns `RG01` through `RG20_25`
```{r seattle_rain_tall}
library(tidyr) 
seattle_rain_tall <- seattle_rain %>% 
  gather(RG01:RG20_25, key = rain_gauge, value = precip_inches) 
seattle_rain_tall
```

## `gather` all columns *except* `Date`
```{r seattle_rain_tall_fix}
library(tidyr) 
seattle_rain_tall <- seattle_rain %>% 
  gather(key = rain_gauge, value = precip_inches, -Date) # Same result as before!
seattle_rain_tall 
```

## Gathering in SQL
```{sql eval=FALSE}
SELECT date, 'rg01' as rain_gauge, rg01 as precip_inches FROM seattle_rain UNION 
SELECT date, 'rg02' as rain_gauge, rg02 as precip_inches FROM seattle_rain UNION
SELECT date, 'rg03' as rain_gauge, rg03 as precip_inches FROM seattle_rain UNION
SELECT date, 'rg04' as rain_gauge, rg04 as precip_inches FROM seattle_rain UNION
SELECT date, 'rg05' as rain_gauge, rg05 as precip_inches FROM seattle_rain UNION
SELECT date, 'rg07' as rain_gauge, rg07 as precip_inches FROM seattle_rain UNION
SELECT date, 'rg08' as rain_gauge, rg08 as precip_inches FROM seattle_rain UNION
SELECT date, 'rg09' as rain_gauge, rg09 as precip_inches FROM seattle_rain UNION
SELECT date, 'rg10_30' as rain_gauge, rg10_30 as precip_inches FROM seattle_rain UNION
SELECT date, 'rg11' as rain_gauge, rg11 as precip_inches FROM seattle_rain UNION
SELECT date, 'rg12' as rain_gauge, rg12 as precip_inches FROM seattle_rain UNION
SELECT date, 'rg14' as rain_gauge, rg14 as precip_inches FROM seattle_rain UNION
SELECT date, 'rg15' as rain_gauge, rg15 as precip_inches FROM seattle_rain UNION
SELECT date, 'rg16' as rain_gauge, rg16 as precip_inches FROM seattle_rain UNION
SELECT date, 'rg17' as rain_gauge, rg17 as precip_inches FROM seattle_rain UNION
SELECT date, 'rg18' as rain_gauge, rg18 as precip_inches FROM seattle_rain UNION
SELECT date, 'rg20_25' as rain_gauge, rg20_25 as precip_inches FROM seattle_rain
ORDER BY rain_gauge, date;
```

## Questions?

### Thank you!