---
title: "Many models with R `tidyverse` tools"
subtitle: "Visualizing Home Sales Data, Garfield County, CO"
author: "Casey Bates"
date: "1/31/2019"
output:
  ioslides_presentation: 
    smaller: true
  beamer_presentation: default
  slidy_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 6, fig.height = 3.7)
```

## Motivation

Utilize `tidyverse` tools and the `broom` package to generate numerous linear models to evaluate how well sale price is explained by square footage for home sales in Garfield County, Colorado. A dataset is publically available on the [Garfield County Assessor website](https://www.garfield-county.com/assessor/comparable-sales.aspx) that contains 2 years of data from summer 2014 through summer 2016.

```{r data_processing, message=FALSE, include=FALSE}
# This chunk includes the data processing steps to wrangle and clean the data
library(tidyverse) # For reading-in, wrangling, visualizing the data, and for pipe operator
library(readxl) # Tidyverse package for reading in Excel files
library(magrittr) # For the '%<>%' pipe operator

# Read-in townhomes and condo sales data
townhomes <- read_xlsx("2017-comparable-sales-condos-townhomes.xlsx") 

# Read-in single family home sales data
single_family <- read_xlsx("2017-comparable-sales-single-family.xlsx") 

# Check column names and formats
glimpse(townhomes) 
glimpse(single_family)
# 'sale_price' column was coerced to numeric and special characters '$' and ',' removed

# Replace spaces in column names with underscores ('_') and make names lowercase
# the package 'magrittr' is used for the pipe operator, this is loaded with the 'tidyverse' library
colnames(townhomes) %<>% str_replace_all("\\s", "_") %<>% tolower() 
colnames(single_family) %<>% str_replace_all("\\s", "_") %<>% tolower()

# Rename column "actual_year_built" as "year_built" for consistency between datasets
townhomes <- townhomes %>% 
  rename(year_built = actual_year_built) 

# Add a "classifcation" column to the single_family dataset and set all values to "Single Family"
single_family <- single_family %>% 
  mutate(classification = "Single Family")

# bind_rows() throws an error b/c 'reception' column is numeric in townhomes dataset; convert to character
townhomes$reception <- as.character(townhomes$reception)

# Combine the datasets into one
# This dataset contains errors which we will see later 
home_sales_errors <- bind_rows(single_family, townhomes)
glimpse(home_sales_errors)
# No new columns were created during the 'bind_rows()' process, indicating all colums align
# Remove single_family and townhomes dataframes because they are no longer needed
rm(single_family, townhomes)

# unique(home_sales_errors$classification) reveals a "Garage Only" type. Drop this.
home_sales_errors <- home_sales_errors %>% filter(classification != "Garage Only") 

# Rename the column "heated area" to "square_feet" for clarity, though technically not correct
home_sales_errors <- home_sales_errors %>% 
  rename(square_feet = heated_area)
```

## Outline

* Part 1: Exploring the dataset with `ggplot2`
  * Tidying wide datasets with `tidyr`
* Part 2: Many models with `purrr` and `broom`

## Part 1: Exploring the dataset with `ggplot2`
### Processing the data

* Import two Excel files: 

  1. single family home sales, and 
  1. condo & townhome sales
* Replace spaces in column names with underscore and make lowercase
* Rename some columns
* Add `classification` column to `single_family` dataset
    * Set all values to "Single Family"
* Use `bind_rows()` to combine the datasets into one
* Remove "Garage Only" observations


## Glimpse of the data
```{r glimpse}
glimpse(home_sales_errors)
```

## `ggplot2` package in R
* Created by Hadley Wickham
* Built on the "Grammar of Graphics" principles
* Core **tidyverse** package
* Every ggplot2 plot has 3 key components:
  * **Data**
  * **Aesthetic mappings** between variables and visuals 
  * Layer(s) to describe how to render each observation (usually created with a **geom** function)

## Basic scatterplot
```{r scatter_location}
ggplot(data = home_sales_errors, aes(x = square_feet, y = sale_price)) +
  geom_point()
```

## Transparency and color by location
```{r scatter_location_color, warning= FALSE}
ggplot(data = home_sales_errors, 
       aes(x = square_feet, y = sale_price, color = location)) +
  geom_point(alpha = 0.5)
```

## Zooming into sales above $1M
```{r scatter_zoom, warning= FALSE}
ggplot(data = home_sales_errors, 
       aes(x = square_feet, y = sale_price, color = location)) +
  geom_point(alpha = 0.5) +
  coord_cartesian(ylim = c(1000000, 3000000))
```

## Add random noise with jitter
```{r scatter_zoom_jitter, warning= FALSE}
ggplot(data = home_sales_errors, 
       aes(x = square_feet, y = sale_price, color = location)) +
  geom_jitter(alpha = 0.5, width = 300, height = 20000) +
  coord_cartesian(ylim = c(1000000, 3000000))
```

```{r, include = FALSE}
# High sale price observations
home_sales_errors %>% arrange(desc(sale_price)) %>% select(c("sale_price", "location", "classification", "bedrooms", "square_feet")) %>% head(30)
```

```{r, include = FALSE}
# Remove erroneous high sale price observations
home_sales_fixed <- home_sales_errors %>% filter(location != "RIFLE" | classification != "Townhome" | sale_price != 2200000)
home_sales_fixed %>% filter(location == "RIFLE" & classification == "Townhome") %>% arrange(desc(sale_price)) %>% head(10)
```

## Linear model: sale price vs. square ft.
```{r lm_overall_bw, warning= FALSE}
ggplot(data = home_sales_errors, aes(x = square_feet, y = sale_price)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) # Method set to lm for the linear model
```

## Linear model: sale price vs. square ft.
```{r lm_overall_fixedbw, warning= FALSE}
ggplot(data = home_sales_fixed, aes(x = square_feet, y = sale_price)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) # Method set to lm for the linear model
```

## Linear model: sale price vs. square feet
```{r lm_overall, warning= FALSE}
ggplot(data = home_sales_errors, 
       aes(x = square_feet, y = sale_price, color = location)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) # Method set to lm for the linear model
```

## Linear model with errors removed
```{r lm_overall_fixed, warning= FALSE}
ggplot(data = home_sales_fixed, 
       aes(x = square_feet, y = sale_price, color = location)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) # Method set to lm for the linear model
```

## Linear model facetted by location
```{r lm_facet, warning= FALSE}
ggplot(home_sales_errors, aes(square_feet, sale_price, color = location)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ location)
```

## Linear model facetted by location
```{r lm_facet_fixed, warning= FALSE}
ggplot(home_sales_fixed, aes(square_feet, sale_price, color = location)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ location)
```

## Boxplots ordered by median sale price
```{r boxplot_fixed, warning = FALSE}
ggplot(home_sales_fixed, 
       aes(x = reorder(location, sale_price, fun = median), 
           y = sale_price, fill = location)) + 
  geom_boxplot() + theme(legend.position = "bottom") +
  labs(x = "Ordered by Median Sale Price", y = "Sale Price")
```

## Boxplots ordered by median sale price
```{r boxplot, warning = FALSE}
ggplot(home_sales_errors, 
       aes(x = reorder(location, sale_price, fun = median), 
           y = sale_price, fill = location)) + 
  geom_boxplot() + theme(legend.position = "bottom") +
  labs(x = "Ordered by Median Sale Price", y = "Sale Price")
```

```{r seattle_rain, include = FALSE}
# Import seattle rain gauge dataset
seattle_rain <- read_csv("Observed_Monthly_Rain_Gauge_Accumulations_Oct_2002_to_May_2017.csv")
```

## Tidying wide datasets
```{r head_rain}
seattle_rain # Display Monthly Rain Gauge Accumulations for Seattle
```

## `gather` columns `RG01` through `RG20_25`
```{r seattle_rain_tall}
library(tidyr) 
seattle_rain_tall <- seattle_rain %>% 
  gather(RG01:RG20_25, key = rain_gauge, value = precip_inches) 
seattle_rain_tall
```

## `gather` all columns *except* `Date`
```{r seattle_rain_tall_fixed}
library(tidyr) 
seattle_rain_tall <- seattle_rain %>% 
  gather(key = rain_gauge, value = precip_inches, -Date) # Same result as before!
seattle_rain_tall 
```

## ...or you could do it in SQL :)
```{sql eval=FALSE}
SELECT date, 'rg01' as rain_gauge, rg01 as precip_inches FROM seattle_rain UNION 
SELECT date, 'rg02' as rain_gauge, rg02 as precip_inches FROM seattle_rain UNION
SELECT date, 'rg03' as rain_gauge, rg03 as precip_inches FROM seattle_rain UNION
SELECT date, 'rg04' as rain_gauge, rg04 as precip_inches FROM seattle_rain UNION
SELECT date, 'rg05' as rain_gauge, rg05 as precip_inches FROM seattle_rain UNION
SELECT date, 'rg07' as rain_gauge, rg07 as precip_inches FROM seattle_rain UNION
SELECT date, 'rg08' as rain_gauge, rg08 as precip_inches FROM seattle_rain UNION
SELECT date, 'rg09' as rain_gauge, rg09 as precip_inches FROM seattle_rain UNION
SELECT date, 'rg10_30' as rain_gauge, rg10_30 as precip_inches FROM seattle_rain UNION
SELECT date, 'rg11' as rain_gauge, rg11 as precip_inches FROM seattle_rain UNION
SELECT date, 'rg12' as rain_gauge, rg12 as precip_inches FROM seattle_rain UNION
SELECT date, 'rg14' as rain_gauge, rg14 as precip_inches FROM seattle_rain UNION
SELECT date, 'rg15' as rain_gauge, rg15 as precip_inches FROM seattle_rain UNION
SELECT date, 'rg16' as rain_gauge, rg16 as precip_inches FROM seattle_rain UNION
SELECT date, 'rg17' as rain_gauge, rg17 as precip_inches FROM seattle_rain UNION
SELECT date, 'rg18' as rain_gauge, rg18 as precip_inches FROM seattle_rain UNION
SELECT date, 'rg20_25' as rain_gauge, rg20_25 as precip_inches FROM seattle_rain
ORDER BY rain_gauge, date;
```

## Part 2: Fit many models with `purrr` and `broom`

* `broom::tidy(model)`
  * Returns 1 row for each coefficient
  * Columns present info about variability or estimates

* `broom::augment(model, data)`
  * Returns 1 row for each row in the data
  * Adds residuals, influence statistics

* `broom::glance(model)`
  * Returns 1 row for each model
  * Each column represents a model summary (quality and/or complexity)

## Workflow with `broom`
`tidyr::nest()` %>% `purrr::map()` %>% `tidyr::unnest()`

### Typical 4-step process
1. `nest()` dataset by categorical variable
1. Fit models to nested lists with `map()`
1. Apply `broom::tidy`, `broom::augment`, and/or `broom::glance` to each nested model
1. `unnest()` to tidy dataframe

*The 4-step process can also be applied to other packages and functions such as `modelr::add_residuals`*

## Why `broom`?
**While model inputs usually require tidy inputs, such attention to detail doesn’t carry over to model outputs. Outputs such as predictions and estimated coefficients aren’t always tidy. This makes it more difficult to combine results from multiple models.** For example, in R, the default representation of model coefficients is not tidy because it does not have an explicit variable that records the variable name for each estimate, they are instead recorded as row names. In R, row names must be unique, so combining coefficients from many models (e.g., from bootstrap resamples, or subgroups) requires workarounds to avoid losing important information. **This knocks you out of the flow of analysis and makes it harder to combine the results from multiple models. I’m not currently aware of any packages that resolve this problem.**

Hadley Wickham 

*Emphasis added by David Robinson, author of the `broom` package, in [this post](http://varianceexplained.org/r/broom-intro/).*

## Linear model for Carbondale
```{r carbondale_lm}
carbondale <- home_sales_fixed %>% 
  filter(location == "CARBONDALE")
model <- lm(sale_price ~ square_feet, data = carbondale)
summary(model)
```

## Transform model output into tidy data frame with `broom::tidy()`
```{r tidy_model}
library(broom)
tidy(model)
```

## `bind_rows` to combine models
```{r two_models}
carbondale <- home_sales_fixed %>% 
  filter(location == "CARBONDALE")
model1 <- lm(sale_price ~ square_feet, data = carbondale)
glenwood <- home_sales_fixed %>% 
  filter(location == "GLENWOOD")
model2 <- lm(sale_price ~ square_feet, data = glenwood)
bind_rows(tidy(model1), tidy(model2))
```

## Step 1: `nest` by location
```{r step_1}
home_sales_fixed_models <- home_sales_fixed %>% 
  nest(-location)
home_sales_fixed_models
```

## Alternate syntax used by Wickham
```{r alt_nest}
home_sales_fixed %>% 
  group_by(location) %>% 
  nest()
```

## Examine the data for New Castle
```{r new_castle}
home_sales_fixed_models$data[[3]]
```

## Unnesting returns to original
```{r unnest}
unnest(home_sales_fixed_models)
```

## Step 2: `map()` to fit lm to each list
```{r step_2}
library(purrr)
home_sales_fixed_models <- home_sales_fixed %>%
  nest(-location) %>% 
  mutate(models = map(data, ~lm(sale_price ~ square_feet, .))) 
  # data has been passed into lm through map function
  # dot "." is used for data in the lm call
home_sales_fixed_models
```

## Examine the model for New Castle
```{r new_castle_v2}
home_sales_fixed_models$models[[3]]
```

## Step 3: Use `map()` to tidy each model
```{r step_3}
home_sales_fixed_models <- home_sales_fixed %>%
  nest(-location) %>% 
  mutate(models = map(data, ~lm(sale_price ~ square_feet, .))) %>% 
  mutate(tidied = map(models, tidy))
home_sales_fixed_models
```

## Examine tidy model for New Castle
```{r new_castle_tidy}
home_sales_fixed_models$tidied[[3]]
```

## Step 4: `unnest` to tidy table of coefficients
```{r step_4}
location_coeffs <- home_sales_fixed %>%
  nest(-location) %>% 
  mutate(models = map(data, ~lm(sale_price ~ square_feet, .))) %>% 
  mutate(tidied = map(models, tidy)) %>% 
  unnest(tidied)
location_coeffs
```

## Location slopes
```{r slopes}
location_slopes <-  location_coeffs %>% 
  filter(term == "square_feet")%>% 
  arrange(estimate)
location_slopes
```

## Easily add `p.adjust()` column
```{r p_adjust}
location_slopes <-  location_coeffs %>% 
  filter(term == "square_feet")%>% 
  mutate(p.adjusted = p.adjust(p.value)) %>% 
  arrange(estimate)
location_slopes
```

## Apply this process/workflow to other packages

Before we move on to `broom:augment` and `broom::glance()`, let's apply this process to `modelr::add_residuals()`.

## Compute residuals for Rifle (errors)
```{r rifle_resids, message = FALSE}
library(modelr) # for add_residuals function
rifle <- filter(home_sales_errors, location == "RIFLE")
rifle_model <- lm(sale_price ~ square_feet, data = rifle)
rifle %>% 
  add_residuals(rifle_model) %>% 
  ggplot(aes(square_feet, resid)) + # new resid column
  geom_hline(yintercept = 0, color = "white", size = 3) +
  geom_line() +
  ggtitle("Remaining pattern")
```

## Compute residuals for Rifle (fixed)
```{r rifle_resids_v2}
rifle <- filter(home_sales_fixed, location == "RIFLE")
rifle_model <- lm(sale_price ~ square_feet, data = rifle)
rifle %>% 
  add_residuals(rifle_model) %>% 
  ggplot(aes(square_feet, resid)) +
  geom_hline(yintercept = 0, color = "white", size = 3) +
  geom_line() +
  ggtitle("Remaining pattern")
```

## `mutate()` [add] a residuals column 
```{r add_resids, message=FALSE}
home_sales_fixed_models <- home_sales_fixed_models %>% 
  # Add 'resid' column to 'data' list columns with mutate 
  # Call add_residuals() with each model-data pair
  mutate(data = map2(data, models, add_residuals)) 
home_sales_fixed_models
```

## Examine residuals for New Castle
```{r new_castle_resids}
home_sales_fixed_models$data[[3]] %>%
  # Select a few columns...
  select(sale_price, square_feet, resid)
```

## `unnest` list columns to plot resids
```{r unnest_resids, warning= FALSE}
resids <- unnest(home_sales_fixed_models, data)
resids %>% 
  ggplot(aes(square_feet, resid)) +
  geom_line(aes(group = location, color=location), alpha = 1 / 3) +
  geom_smooth(se = FALSE)
```

## Or facet by location
```{r resids_facet_fixed, warning= FALSE}
resids %>% 
  ggplot(aes(square_feet, resid, color = location)) +
  geom_line(alpha = 1 / 3) +
  facet_wrap(~ location)
```

## Pipe it all together!
```{r resids facet, warning= FALSE}
home_sales_errors %>% 
  nest(-location) %>% # Step 1
  mutate(models = map(data, ~lm(sale_price ~ square_feet, .))) %>% # Step 2
  mutate(data = map2(data, models, add_residuals)) %>% # Step 3
  unnest(data) %>% #Step 4
  ggplot(aes(square_feet, resid, color = location)) + #Step 5 - plot!
  geom_line(alpha = 1 / 3) +
  facet_wrap(~ location)
```

## `broom::augment()` the results 
```{r augment, message=FALSE}
home_sales_fixed_models <- home_sales_fixed_models %>% 
  mutate(augmented = map(models, broom::augment))
home_sales_fixed_models
```

## `glimpse` at the list column `augmented`
```{r augment_glimpse}
augment_results <- home_sales_fixed_models %>%
  unnest(augmented)
glimpse(augment_results)
```

## Plot the residuals from `augmented`
```{r augment_resids_plot}
augment_results %>%  
ggplot(aes(square_feet, .resid, color = location)) +
  geom_line(alpha = 1 / 3) +
  facet_wrap(~ location)
```

## `broom::glance()` at model summaries
```{r glance}
home_sales_fixed_models <- home_sales_fixed_models %>% 
  mutate(glanced = map(models, broom::glance))
home_sales_fixed_models
```

## `glance()` results
```{r glance_listcols}
glance_results <- home_sales_fixed_models %>% 
  unnest(glanced)
glance_results
```

## `glance()` results without list columns
```{r glance_listcols2}
glance_results <- home_sales_fixed_models %>% 
  # Supress nested columns with ".drop = TRUE"
  unnest(glanced, .drop = TRUE)
glance_results
```

## Resources
* [`broom` and `dplyr` vignette](https://cran.r-project.org/web/packages/broom/vignettes/broom_and_dplyr.html)
* [`broom` intro by David Robinson](http://varianceexplained.org/r/broom-intro/)
* [R for Data Science by Hadley Wickham](https://r4ds.had.co.nz/)

## Questions?

casey.bates@erm.com

Thank you!